---
title: "Good to Great: Custom Themes in ggplot2"
subtitle: ""
author: "Meghan Hall"
institute: "Virtual TidyTuesday UseR Meetup"
date: "March 24, 2020"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false

---

# Me in 15 seconds

--

- Day job: data manager in higher education (I use R a lot)

--

- Also a hockey nerd! (I use R a lot there too)

--

- R makes my work so much easier, so I love helping people learn

--

- Find me on Twitter [@MeghanMHall](https://twitter.com/MeghanMHall)

- Check out [Hockey Graphs](https://hockey-graphs.com/)

---
class: inverse, center, middle

# Onto Themes!

---

# Why Use a Custom Theme?

--

- It helps brand your work! Whether you're working for yourself or for a company.

--

- It shows that you know R and are willing to put in a little effort for a custom look. 

--

- Most importantly: It's not as hard as it seems, I swear.

---

# Level One

A simple plot with **ggplot2**:

```{r plot-label1, fig.show = 'hide'}
library(datasets)
library(tidyverse)

ggplot(chickwts) +
  geom_bar(aes(x = feed, y = weight, fill = feed), 
         position = "dodge", stat = "summary", fun.y = "mean") +
  labs(
    y = "Average Weight",
    x = "Feed",
    fill = NULL,
    title = "Average Chicken Weight by Feed Type",
    subtitle = "What's the unit of measure? Who knows!",
    caption = "From the chickwts dataset"
  )
```

---

# Level One

.center[

```{r ref.label = 'plot-label1', echo = FALSE}
```

]

---

# Level Two

Add in a **built-in theme** and some **custom colors**:

```{r plot-label2, fig.show = 'hide'}
library(datasets)
library(tidyverse)

ggplot(chickwts) +
  geom_bar(aes(x = feed, y = weight, fill = feed), 
         position = "dodge", stat = "summary", fun.y = "mean") +
  labs(
    y = "Average Weight",
    x = "Feed",
    fill = NULL,
    title = "Average Chicken Weight by Feed Type",
    subtitle = "What's the unit of measure? Who knows!",
    caption = "From the chickwts dataset"
  ) +
  scale_fill_manual(values = c("#2B4162", "#C5CBD4", "#385F71", 
                               "#D7B377", "#F5F0F6", "#8F754F")) +
  theme_minimal()
```

---

# Level Two

.center[

```{r ref.label = 'plot-label2', echo = FALSE}
```

]

---

# Built-In Themes

- theme_gray()

- theme_bw()

- theme_linedraw()

- theme_light()

- theme_dark()

- theme_minimal()

- theme_classic()


---

# Level Three

Use your own **custom theme** and some **custom color palettes**:

```{r plot-label3, eval=FALSE}
library(datasets)
library(tidyverse)

ggplot(chickwts) +
  geom_bar(aes(x = feed, y = weight, fill = feed), 
         position = "dodge", stat = "summary", fun.y = "mean") +
  labs(
    y = "Average Weight",
    x = "Feed",
    fill = NULL,
    title = "Average Chicken Weight by Feed Type",
    subtitle = "What's the unit of measure? Who knows!",
    caption = "From the chickwts dataset"
  ) +
  scale_fill_manual(values = my_colors) +
  meg_theme()
```

---

# Level Three

.center[

```{r plot-label3-out, ref.label="plot-label3", echo=FALSE}
```

]

---

# Custom Color Palettes

- If you find yourself constantly copying and pasting the same hex codes, it might be time to make a custom color palette!

```{eval=FALSE}
my_colors <- c("#2B4162", "#C5CBD4", "#385F71", 
               "#D7B377", "#F5F0F6", "#8F754F")

+ scale_fill_manual(values = my_colors)

+ scale_color_manual(values = my_colors)
```

--

- Can also use the ColorBrewer scales that are part of ggplot2 (various palettes available, can view with ?scale_fill_brewer).

- Always consider colorblindness: try viridis in ggplot2 or find a palette generator with a colorblind option (I like [coolors.co](coolors.co)).

```{eval=FALSE}
+ scale_fill_brewer(palette = "Pastel1")

+ scale_fill_viridis_d()
```

---

# Making Your Own Theme

- In my experience, it was easier to start from an existing theme.

```{eval=FALSE}
  theme_linedraw(base_size=11, base_family="Courier") %+replace% 
    theme(
      axis.ticks = element_blank(),
      panel.grid.minor = element_blank()
    )
```

--

- [This ggplot2 reference page](https://ggplot2.tidyverse.org/reference/theme.html) is your bible, it lists all the various arguments to the theme() function.

--

Just a few of the arguments I use in meg_theme():

```{eval=FALSE}
plot.title = element_text(size = 13, hjust = 0, vjust = 0.5, 
                margin = margin(b = 0.2, unit = "cm")),
plot.subtitle = element_text(size = 9, hjust = 0, vjust = 0.5, 
                   margin = margin(b = 0.2, unit = "cm")),
plot.caption = element_text(size = 7, hjust = 1, face = "italic", 
                   margin = margin(t = 0.1, unit = "cm")
```

---

# Making Your Own Theme

How do you know what all the options are?

--

```{r comment='#'}
theme_deets <- theme_linedraw()

theme_deets$plot.caption
```

---

# Using Your Own Theme

```{r plot-label4, eval=FALSE}
our_new_theme <- function () { 
  theme_linedraw(base_size=11, base_family="Courier") %+replace% 
    theme(
      axis.ticks = element_blank(),
      panel.grid.minor = element_blank()
    )}

ggplot(chickwts) +
  geom_bar(aes(x = feed, y = weight, fill = feed), 
         position = "dodge", stat = "summary", fun.y = "mean") +
  labs(
    y = "Average Weight",
    x = "Feed",
    fill = NULL,
    title = "Average Chicken Weight by Feed Type",
    subtitle = "What's the unit of measure? Who knows!",
    caption = "From the chickwts dataset"
  ) +
  our_new_theme()
```

---

# Using Your Own Theme

.center[

```{r plot-label4-out, ref.label="plot-label4", echo=FALSE}
```

]

---

class: inverse, center, middle

# Thanks! Have fun theme building!

Slides created with the R package [xaringan](https://github.com/yihui/xaringan).

I'm on Twitter [@MeghanMHall](https://twitter.com/MeghanMHall).
