---
title: Creating Custom Color Palettes with ggplot2
date: '2022-10-13'
summary: 'How to define custom colors and color palettes in R and create your own scale_color and scale_fill functions to use in ggplot2.'
categories: ["R", "ggplot2", "data viz"]
---

```{r options, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE)

library(tidyverse)
library(palmerpenguins)

penguins <- palmerpenguins::penguins

theme_here <- function () { 
  theme_linedraw() %+replace% 
    theme(
      panel.background = element_rect(fill = "transparent", color = NA),
      plot.background = element_rect(fill = "transparent", color = NA),
      legend.background = element_rect(fill = "transparent", color = NA),
      legend.key = element_rect(fill = "transparent", color = NA)
    )
}
```

## Why custom palettes?

One of the perks of making data visualizations in `ggplot2` is that nearly everything is easily customizable, meaning your plots can stick to a common style with the fonts and elements specified by your organization's or your own personal style guide (hopefully you have a custom theme to make this easier!). When it comes to using colors in data visualizations, if you find yourself typing in the same hex codes often enough that you start to memorize them---speaking from experience here---it's probably time to codify those colors and palettes so you can reference them more easily when creating plots. The examples below will show you how to specify colors and palettes and then use those palettes to create your own `scale_color` and `scale_fill` functions for use in plots, for both discrete and continuous data.

[This blog post](https://drsimonj.svbtle.com/creating-corporate-colour-palettes-for-ggplot2) by [Simon Jackson](https://twitter.com/drsimonj) and [this one](https://www.garrickadenbuie.com/blog/custom-discrete-color-scales-for-ggplot2/) by [Garrick Aden-Buie](https://twitter.com/grrrck) were both incredibly helpful in helping me figure out how to do this. I somewhat combined their approaches to get this method that worked for my needs, and hopefully either this method or one of theirs is helpful for you!

## Defining custom colors and palettes

```{r setup, eval=FALSE}
library(tidyverse)
library(palmerpenguins)

penguins <- palmerpenguins::penguins
```

These examples use the `penguins` data set from the wonderful [`palmerpenguins`](https://allisonhorst.github.io/palmerpenguins/) package. Let's imagine that you work for Penguin Corp and regularly create reports and visualizations with various penguin-related data. Penguin Corp has a style guide that specifies the colors and palettes that should be used, and while you can always specify the colors manually in `ggplot2` visualizations, it's much easier to define the colors and palettes and then reference them through your own `scale_fill` and `scale_color` functions. (These are all functions that can be dropped easily into an internal package!)

First, start by defining the various colors you need.

```{r color}
penguin_corp_color <- function(...) {

  penguin_corp_colors <- c(
    `pink`     = "#F7B1AB",
    `lavender` = "#807182",
    `gray`     = "#E3DDDD",
    `brown`    = "#A45C3D",
    `purple`   = "#1C0221")

  cols <- c(...)

  if (is.null(cols))
    return (penguin_corp_colors)

  penguin_corp_colors[cols]
}
```

That function can be used to show the definition of a specific color:

```{r color-example}
penguin_corp_color("lavender")
```

That `penguin_corp_color` function then becomes the base of the `penguin_corp_palette` function below, where those defined colors are combined into palettes. Your organization might have primary and secondary palettes, or palettes designed for specific uses, but here we'll define a `main` palette as well as a `highlight` palette for when we want just two colors.

```{r palette}
penguin_corp_palette <- function(palette = "main", ...) {

  penguin_corp_palettes <- list(
    `main` = penguin_corp_color("lavender", "gray", "pink", "brown"),
    
    `highlight` = penguin_corp_color("purple", "gray")
  )

  penguin_corp_palettes[[palette]]

}
```

```{r palette-example}
penguin_corp_palette("main")
```

The `show_col` function from the `scales` package is a nifty way to showcase all the colors available in a given palette:

```{r palette-demo, out.width="50%", fig.align="center"}
scales::show_col(penguin_corp_palette("main"), cex_label = 2)
```

## Creating your own scale_color and scale_fill functions

Having the colors and palettes defined is a great first step, but you can go even further and apply those into your own `scale_fill` and `scale_color` functions. The first step is a helper function, here called `palette_gen`. This function has two arguments, the name of the palette ("main" will be our default) and the direction (so you can flip the scale if necessary), and essentially creates *another* function that will be used in the `scale_fill` and `scale_color` functions. (The `n` there refers to the number of colors that would be needed for a particular plot.)

```{r pal-gen}
palette_gen <- function(palette = "main", direction = 1) {

  function(n) {

    if (n > length(penguin_corp_palette(palette)))
      warning("Not enough colors in this palette!")

    else {

      all_colors <- penguin_corp_palette(palette)

      all_colors <- unname(unlist(all_colors))

      all_colors <- if (direction >= 0) all_colors else rev(all_colors)

      color_list <- all_colors[1:n]

    }
  }
}
```

The function above is for discrete color scales. If you also want to use continuous color scales, the function below uses the existing `colorRampPalette` function to interpolate the necessary colors between the ones you have chosen in your palette.

```{r}
palette_gen_c <- function(palette = "main", direction = 1, ...) {

  pal <- penguin_corp_palette(palette)

  pal <- if (direction >= 0) pal else rev(pal)

  colorRampPalette(pal, ...)

}
```

With that helper function created, you can write the actual functions to be used with `ggplot2`. I've called mine `scale_fill_penguin`, which takes the same two arguments as before: `palette` and `direction`.

```{r scale_fill}
scale_fill_penguin <- function(palette = "main", direction = 1, ...) {

  ggplot2::discrete_scale(
    "fill", "penguin",
    palette_gen(palette, direction),
    ...
  )
}
```

You can use the same syntax for `scale_color`. (Fun fact: I learned from Garrick's post that the `ggplot2` convention is to create a `scale_colour` function and then replicate it as `scale_color`.)

```{r scale_color}
scale_colour_penguin <- function(palette = "main", direction = 1, ...) {

  ggplot2::discrete_scale(
    "colour", "penguin",
    palette_gen(palette, direction),
    ...
  )
}

scale_color_penguin <- scale_colour_penguin
```

Again, those are for discrete color scales. If you need a continuous scale, use a function like this:

```{r scale_color_c}
scale_color_penguin_c <- function(palette = "main", direction = 1, ...) {

  pal <- palette_gen_c(palette = palette, direction = direction)

  scale_color_gradientn(colors = pal(256), ...)

}
```

## Using these colors in plots

Let's look at some examples to see the various ways these functions can be used to customize colors with `ggplot2`. Below is a very simple bar chart using the `palmerpenguins::penguins` data set.

```{r plot-basic, fig.align="center"}
penguins %>% 
  count(species) %>% 
  ggplot(aes(x = species, y = n)) +
  geom_bar(stat = "identity") +
  labs(title = "The count of each species in the palmerpenguins data set") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  theme_here() +
  theme(axis.ticks = element_blank(),
        axis.title = element_blank(),
        panel.grid.major.x = element_blank())
```

The new `penguin_corp_color` function can be used to specifically define a color. In the plot below, `penguin_corp_color("lavender")` is the `fill` argument in the `geom_bar` layer to make all of the bars that specific shade of lavender.

```{r plot-single, fig.align="center"}
penguins %>% 
  count(species) %>% 
  ggplot(aes(x = species, y = n)) +
  geom_bar(stat = "identity", fill = penguin_corp_color("lavender")) +
  labs(title = "The count of each species in the palmerpenguins data set") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  theme_here() +
  theme(axis.ticks = element_blank(),
        axis.title = element_blank(),
        panel.grid.major.x = element_blank())
```

We could also add `fill = species` to the `aes` layer (meaning that each species will have its own color) and then use `scale_fill_penguin(palette = "main")` to automatically apply our "main" color palette.

```{r plot-palette, fig.align="center"}
penguins %>% 
  count(species) %>% 
  ggplot(aes(x = species, y = n, fill = species)) +
  geom_bar(stat = "identity") +
  scale_fill_penguin(palette = "main") +
  labs(title = "The count of each species in the palmerpenguins data set") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  theme_here() +
  theme(axis.ticks = element_blank(),
        axis.title = element_blank(),
        legend.position = "none",
        panel.grid.major.x = element_blank())
```

Or if you want to use the custom colors but *not* a specific palette, add a `scale_fill_manual` layer and specify the values using the `penguin_corp_color` function.

```{r plot-multiple, fig.align="center"}
penguins %>% 
  count(species) %>% 
  ggplot(aes(x = species, y = n, fill = species)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = 
                      unname(c(penguin_corp_color("brown","gray","purple")))) +
  labs(title = "The count of each species in the palmerpenguins data set") +
  scale_y_continuous(expand = expansion(mult = c(0, 0.1))) +
  theme_here() +
  theme(axis.ticks = element_blank(),
        axis.title = element_blank(),
        legend.position = "none",
        panel.grid.major.x = element_blank())
```

Lastly, below is an example of a continuous color scale, with our `scale_color_penguin_c` function specifying the palette to be used.

```{r plot-continuous, fig.align="center"}
penguins |> 
  filter(!is.na(sex)) |> 
  ggplot(aes(x = sex, y = body_mass_g, color = bill_depth_mm)) +
  geom_jitter(size = 3, width = 0.3) +
  scale_color_penguin_c(palette = "highlight", direction = -1) +
  labs(title = "Bill depth and body mass by sex",
       y = "Body mass (g)",
       color = "Bill depth (mm)") +
  theme_here() +
  theme(axis.ticks = element_blank(),
        axis.title.x = element_blank(),
        panel.grid.major.x = element_blank())
```