---
format: 
  pdf: 
    mainfont: "Avenir"
    sansfont: "Avenir"
    geometry:
      - top=0.75in
      - right=0.75in
      - bottom=0.75in
      - left=0.75in
      - heightrounded
    number-sections: true
    include-in-header: 
      - "penguin-header.tex"
    toc: false
editor: source
execute:
  warning: false
  echo: false
params:
  species: 'Adelie'
  year: 2009
---
\sffamily
\thispagestyle{empty}
\color{`r params$species`}
\includegraphics[width=4cm]{penguins.png}\hfill\large\texttt{palmerpenguins}\textbf{: `r params$species` penguins in `r params$year`}\normalsize
\vspace{-0.2in}
\rule{17.78cm}{0.5mm}

```{r setup}
knitr::opts_chunk$set(echo = FALSE)
options(knitr.kable.NA = '')

library(tidyverse)
library(palmerpenguins)
library(glue)
library(kableExtra)

penguins <- penguins

color <- case_when(params$species == "Adelie" ~ "#FF6600",
                   params$species == "Gentoo" ~ "#047075",
                   params$species == "Chinstrap" ~ "#C25BCC")
```

\color{black}

The `r params$species` penguins can be found on the island(s) of: `r glue_collapse(penguins |> filter(species == params$species) |> count(island) |> pluck(1), sep = ", ", last = ", and ")`. In `r params$year` there were `r nrow(penguins |> filter(species == params$species & year == params$year))` penguins measured.

\color{`r params$species`}
# Mass

\color{black}
```{r table}
penguins %>%  
  filter(year == params$year & !is.na(sex)) %>% 
  group_by(species, sex) |> 
  summarize(mass = round(mean(body_mass_g, na.rm = TRUE), 1)) |> 
  kbl(booktabs = T, escape = F, longtable = T, linesep = "",
      col.names = c("Species", "Sex", "Avg Mass (g)"),
      align = c(rep("l", 2), "c")) %>% 
  kable_classic(full_width = F) %>% 
  row_spec(0, bold = TRUE) %>%
  row_spec(case_when(params$species == "Adelie" ~ c(1,2),
                     params$species == "Chinstrap" ~ c(3,4),
                     params$species == "Gentoo" ~ c(5,6)), 
           color = color, bold = TRUE) %>%
  kableExtra::kable_styling(latex_options = "hold_position")
```

\color{`r params$species`}
# Bills

```{r scatter}
#| fig-height: 3
#| fig-align: center
library(showtext)

font_add_google(name = "Prompt", 
                family = "prompt")
showtext_auto()

penguin_theme <- function () { 
  theme_linedraw(base_size=11, base_family="prompt") %+replace% 
    theme(
      panel.background  = element_blank(),
      plot.background = element_rect(fill = "transparent", color = NA), 
      axis.ticks = element_blank(),
      panel.grid.major = element_line(color = "grey90", size = 0.3), 
      panel.grid.minor = element_blank(),
      axis.text.x = element_text(size = 13),
      axis.text.y = element_text(size = 13)
    )
}

penguin_plot <- penguins |> 
  filter(year == params$year) |> 
  mutate(selected = species == params$species)

penguin_plot |> 
  ggplot(aes(x = bill_length_mm, y = bill_depth_mm, color = selected)) + 
  scale_color_manual(values = c("grey",color), guide = "none") +
  geom_point(size = 3) +
  labs(x = "Bill length in mm", y = "Bill depth in mm") +
  penguin_theme()
```

\vfill
\textcolor{`r params$species`}{\rule{17.78cm}{0.5mm}}

\center\small\emph{Created `r format(Sys.time(), '%F')`. Artwork by @allison\_horst.}
